---
title: "INLA and R-INLA"
subtitle: "Bayesian Modeling for Environmental Health Workshop"
author: "Robbie M. Parks"
date: "August 20 2026"
format: html
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
library(tidyverse)
library(INLA)
library(nimble)
library(posterior)
library(sf)
library(spdep)
library(hrbrthemes)
library(colorspace)
library(patchwork)
library(knitr)

extrafont::loadfonts()
theme_set(theme_ipsum())

set.seed(2)
```

## Goal of this computing lab session

The lecture made two claims that this lab puts your hands on. First, that a
Laplace approximation is nothing more than a mode and a curvature. Second,
that INLA works by scanning a small number of hyperparameter combinations,
solving the latent field at each, and taking a weighted sum.

We will build both of those from scratch in a few lines of R, then watch
`R-INLA` do the same thing on the Italian mortality data from the
hierarchical modelling session.

## What's going to happen in this lab session?

1. Build a Laplace approximation by hand, and see where it breaks;
2. Build the *integrated* version by hand, and see what integrating buys;
3. Refit the full pooling, no pooling and partial pooling models in `INLA`;
4. Look inside the fitted object at the hyperparameter combinations `INLA`
   actually used;
5. Vary `int.strategy` and `strategy` and watch the accuracy–speed trade-off;
6. Inspect the precision matrix $\boldsymbol{Q}$ and its blanks;
7. Extend to a spatially structured `bym2` model, then a spatiotemporal one;
8. Compare against `NIMBLE`.

## Introduction

In the hierarchical modelling session we estimated the standardised mortality
ratio (SMR) for the provinces of Italy in September 2018 using three models
(full pooling, no pooling and partial pooling), fitted with MCMC in `NIMBLE`.
That workflow required initial values, burn-in, convergence checks, trace
plots and $\hat{R}$.

All three are **latent Gaussian models**, and they satisfy the two conditions
from the lecture:

1. the latent field (intercept, fixed effects, province effects) is Gaussian,
   and its links are local;
2. there are at most **one** hyperparameter, the precision of the province
   effects.

That is why `INLA` can fit them deterministically.

::: aside
`INLA` is not on CRAN. If you have not installed it yet:

```r
install.packages(
  "INLA",
  repos = c(getOption("repos"),
            INLA = "https://inla.r-inla-download.org/R/stable"),
  dep = TRUE
)
```
:::

# Part 1: the two approximations, by hand

Before touching any data, we build the two approximations ourselves.

## The Laplace approximation

Recall from the lecture: expand $\log f(x)$ about its mode, where the first
derivative vanishes, and keep the quadratic term. Exponentiating gives a
Gaussian. The whole approximation is **two numbers**: the mode $x^{*}$, and
the curvature $H = -\log f''(x^{*})$.

We will approximate a $\text{Gamma}(5, 1)$, which is skewed but has a known answer to check against.

```{r}
shape <- 5; rate <- 1

# log density, up to a constant
lf <- function(x) (shape - 1) * log(x) - rate * x

# 1. the mode
mode <- optimise(lf, c(1e-6, 50), maximum = TRUE)$maximum

# 2. the curvature there:  -f''(x) = (shape - 1) / x^2
prec <- (shape - 1) / mode^2
sdev <- 1 / sqrt(prec)

tibble(
  quantity = c("mode x*", "curvature H", "implied sd", "true mean", "true sd"),
  value    = c(mode, prec, sdev, shape / rate, sqrt(shape) / rate)
) |> kable(digits = 3)
```

The approximation is $N(4, 2^2)$. Plot it against the truth:

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
par(mar = c(4, 4, 1, 1))
curve(dgamma(x, shape, rate), -2, 14, n = 400, lwd = 2.5, bty = "n",
      xlab = "x", ylab = "density")
curve(dnorm(x, mode, sdev), add = TRUE, n = 400, lwd = 2.5,
      lty = 2, col = "#c0392b")
abline(v = mode, lty = 3, col = "grey60")
abline(h = 0, col = "grey80")
legend("topright", bty = "n", lwd = 2.5, lty = c(1, 2),
       col = c("black", "#c0392b"),
       legend = c("Gamma(5, 1)", "Laplace approximation"))
```

The peak is matched exactly, which is all the method promises. But the target leans right, so its **mean** (5) sits above its **mode**
(4), and a symmetric curve cannot follow. The Gaussian also has visible
density below zero, where a gamma has none.

Laplace also approximates the *normalising constant*, which is what `INLA`
uses to score each hyperparameter combination:

```{r}
tibble(
  quantity = c("Laplace estimate", "exact (Gamma function)"),
  value    = c(exp(lf(mode)) * sqrt(2 * pi / prec), gamma(shape))
) |> kable(digits = 3)
```

The estimate is within about 2% of the exact value.

::: {.callout-tip}
## Exercise 1

`shape` is a dial. Re-run the plot with `shape <- 50` and then `shape <- 2`.

- Which gives the better approximation, and why?
- What does this tell you about models with many observations per latent
  element versus few? (Think binary outcomes with one observation per random
  effect.)
:::

## Integrating over a hyperparameter

Laplace alone would stop at the best hyperparameter value. The **integrated**
part of INLA is the refusal to do that.

Take a case with a known exact answer: $x \mid \theta \sim N(0, 1/\theta)$
with $\theta \sim \text{Gamma}(2, 2)$, so $\theta$ is a precision, exactly as
in our models. The exact marginal for $x$ is a $t_4$.

```{r}
a <- 2; b <- 2; K <- 7

# 1. scan theta — a grid on the log scale, as INLA does
theta <- exp(seq(log(0.03), log(12), length.out = K))
w     <- dgamma(theta, a, b) * theta      # Jacobian for the log scale
w     <- w / sum(w)                       # plausibility weights

# 2. the conditional for x at each value of theta
x    <- seq(-7, 7, length.out = 400)
cond <- sapply(theta, \(t) dnorm(x, 0, 1 / sqrt(t)))

# 3. integrate: the weighted sum
marg <- as.vector(cond %*% w)
```

Three stages, three lines. Everything `INLA` does to a 168,000-element latent
field has this shape.

```{r}
#| fig-width: 10
#| fig-height: 3.6
#| fig-align: center
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1), cex = 0.95)

tt <- exp(seq(log(0.03), log(12), length.out = 300))
plot(tt, dgamma(tt, a, b), type = "l", log = "x", lwd = 2.5, bty = "n",
     xlab = expression(theta), ylab = "density",
     main = "1. scan the hyperparameter")
points(theta, dgamma(theta, a, b), pch = 19, col = "#c0392b",
       cex = 0.5 + 4 * w)

matplot(x, cond, type = "l", lty = 1, bty = "n", lwd = 1 + 7 * w,
        col = adjustcolor("#2980b9", alpha.f = 0.75),
        xlab = "x", ylab = "density", main = "2. one conditional each")

plot(x, dt(x, 4), type = "l", lwd = 4, col = "grey65", bty = "n",
     xlab = "x", ylab = "density", main = "3. the weighted sum")
lines(x, marg, lwd = 2, lty = 2, col = "#16a085")
lines(x, dnorm(x, 0, 1 / sqrt((a - 1) / b)), lwd = 2, lty = 3, col = "#c0392b")
legend("topright", bty = "n", lwd = c(4, 2, 2), lty = c(1, 2, 3), cex = 0.9,
       col = c("grey65", "#16a085", "#c0392b"),
       legend = c("truth", "integrated", "fixed theta"))
```

Middle panel is the whole idea in miniature: every blue curve is a *Gaussian*,
none of them is the answer, and their weighted sum is a $t$, a shape none of
the ingredients had.

```{r}
eval_at <- c(0, 3, 6)
tibble(
  x                = eval_at,
  truth            = dt(eval_at, 4),
  integrated       = sapply(eval_at, \(xi) sum(w * dnorm(xi, 0, 1 / sqrt(theta)))),
  `fixed theta`    = dnorm(eval_at, 0, 1 / sqrt((a - 1) / b))
) |> kable(digits = 5)
```

Seven points recover the exact answer to three decimals. Fixing $\theta$ at
its mode gets the peak wrong and the far tail wrong by a factor of about
thirty. The tail is where interval endpoints live, and this is
what `int.strategy = "eb"` does, and precisely what it costs.

::: {.callout-tip}
## Exercise 2

- Reduce `K` to 3 and then 2. How few points can you get away with here?
- Replace the log-scale grid with a linear one,
  `theta <- seq(0.03, 12, length.out = K)`, and drop the Jacobian. What
  happens in the tails, and why does `INLA` work on the log scale?
:::

# Part 2: the Italian mortality models

## Exploratory data analysis

```{r}
data <- read_rds(here("data", "italy", "italy_mortality.rds"))
glimpse(data)
```

As in the hierarchical modelling lab, we collapse the time dimension and focus
on September 2018.

```{r}
data <- data |>
  mutate(provincia_id = data |> group_by(SIGLA) |> group_indices()) |>
  filter(year == 2018) |>
  filter(month == 9) |>
  arrange(SIGLA)
```

And load the shapefile for mapping (and, later, for the adjacency structure).

```{r}
shp_italy <- read_rds(here("data", "italy", "italy_shp.rds")) |> arrange(SIGLA)
```

A quick reminder of what the raw SMR looks like.

```{r}
data |>
  ggplot(aes(x = deaths / expected)) +
  geom_histogram()
```

## Full pooling model

Priors
$$
\alpha \sim N(0, 5)
$$

Likelihood
$$
\begin{split}
y_i &\sim \text{Pois}(\mu_i) \quad i = 1,..., N \\
\log(\mu_i) &= \log(E_i) + \alpha
\end{split}
$$

In `INLA` this is one call. Some things to notice compared to the `NIMBLE`
version:

- The offset goes in through `E =`, on the **natural** scale, not logged;
  `INLA` takes the log for us.
- There is no `f()` term, so the hyperparameter vector $\boldsymbol{\theta}$
  is **empty**. There is nothing to integrate over, and Part 1's second stage
  does not happen at all.
- We match the $N(0, 5)$ prior on $\alpha$ via `control.fixed`, remember
  `INLA` parameterises Gaussians by **precision**, so $\text{sd} = 5$ becomes
  $\text{prec} = 1/25 = 0.04$.

```{r}
t_full_inla <- system.time(
  m_pooled <- inla(
    deaths ~ 1,
    family = "poisson",
    E      = expected,
    data   = data,
    control.fixed   = list(mean.intercept = 0, prec.intercept = 0.04),
    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
  )
)

t_full_inla
summary(m_pooled)
```

No initial values, iteration counts or burn-in settings. The fixed-effect
summary gives the posterior of $\alpha$ directly:

```{r}
m_pooled$summary.fixed
```

::: aside
The `kld` column is the Kullback-Leibler divergence between the Gaussian and
the simplified Laplace marginal for that parameter, a direct measure of how
much the skew correction from Part 1 mattered. Values much above $\sim 0.01$
suggest the approximation is straining. Here it should be near zero.
:::

Let's map the (single, national) SMR, working with the full posterior marginal
rather than point summaries. `inla.tmarginal()` transforms a marginal through
a function, here $\exp$, and `inla.zmarginal()` summarises the result.

```{r}
smr_marg_pooled <- inla.tmarginal(exp, m_pooled$marginals.fixed$`(Intercept)`)
inla.zmarginal(smr_marg_pooled)

SMR_full_pooling <- inla.emarginal(function(x) x, smr_marg_pooled)

p_full <- shp_italy |>
  mutate(SMR = SMR_full_pooling) |>
  ggplot(aes(fill = SMR)) +
  geom_sf(colour = "white") +
  scale_fill_continuous_sequential(palette = "Reds") +
  theme_void()

p_full
```

Still pretty boring, eh?

## No pooling model

Each province gets its own fixed effect, with no borrowing of strength.

Priors
$$
\alpha_j \sim N(0, 1) \quad j = 1,..., N_p
$$

Likelihood
$$
\begin{split}
y_i &\sim \text{Pois}(\mu_i) \\
\log(\mu_i) &= \log(E_i) + \alpha_{j[i]}
\end{split}
$$

Province enters as a **fixed effect**. It still lives in the latent
field, a vague Gaussian prior is still a Gaussian prior, but because there
is no `f()` term there is still no hyperparameter.

```{r}
t_no_inla <- system.time(
  m_unpooled <- inla(
    deaths ~ -1 + factor(provincia_id),
    family = "poisson",
    E      = expected,
    data   = data,
    control.fixed   = list(prec = 1),
    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
  )
)

t_no_inla

SMR_no_pooling <- exp(m_unpooled$summary.fixed$`0.5quant`)

p_no <- shp_italy |>
  mutate(SMR = SMR_no_pooling) |>
  ggplot(aes(fill = SMR)) +
  geom_sf(colour = "white") +
  scale_fill_continuous_sequential(palette = "Reds") +
  theme_void()

p_no
```

## Partial pooling model

Now the hierarchical model: a common intercept plus an IID random effect for
each province, whose spread $\sigma_p$ is estimated from the data.

Priors
$$
\begin{split}
\alpha &\sim N(0,5), \\
\theta_j &\sim N(0, \sigma^2_p) \quad j = 1,..., N_p
\end{split}
$$

Likelihood
$$
\begin{split}
y_i &\sim \text{Pois}(\mu_i) \quad i = 1,..., N \\
\log(\mu_i) &= \log(E_i) + \alpha + \theta_{j[i]}
\end{split}
$$

This is the first model with a hyperparameter, so it is the first model where
Part 1's second stage actually runs.

Instead of the half-normal prior on $\sigma_p$ we used in `NIMBLE`, we use a
**penalised complexity (PC) prior** on the precision. The parameterisation
`param = c(1, 0.01)` reads as $P(\sigma_p > 1) = 0.01$, we consider it
unlikely that the between-province SD exceeds 1 on the log scale.

The `f()` term needs an integer or factor **index**; `provincia_id` is one.

```{r}
pc_prec <- list(prec = list(prior = "pc.prec", param = c(1, 0.01)))

t_partial_inla <- system.time(
  m_partial <- inla(
    deaths ~ 1 + f(provincia_id, model = "iid", hyper = pc_prec),
    family = "poisson",
    E      = expected,
    data   = data,
    control.fixed   = list(mean.intercept = 0, prec.intercept = 0.04),
    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
  )
)

t_partial_inla
summary(m_partial)
```

One `f()` term is the entire difference from the pooled model.

### What INLA actually did

This is the part the lecture described and you built by hand in Part 1. With
`config = TRUE` we can look at the hyperparameter combinations that were
evaluated:

```{r}
n_config <- m_partial$misc$configs$nconfig
n_config
```

That is the number of red dots in panel 1 of your Part 1 figure. For each one,
`INLA` solved the latent field once. Here are the values and their weights:

```{r}
configs <- tibble(
  k          = seq_len(n_config),
  log_theta  = map_dbl(m_partial$misc$configs$config, ~ .x$theta[1]),
  log_post   = map_dbl(m_partial$misc$configs$config, ~ .x$log.posterior)
) |>
  mutate(
    precision = exp(log_theta),
    sigma_p   = 1 / sqrt(precision),
    weight    = exp(log_post - max(log_post)),
    weight    = weight / sum(weight)
  ) |>
  arrange(log_theta)

configs |> select(k, precision, sigma_p, weight) |> kable(digits = 4)
```

Note the grid is laid out on **log precision**, not precision, exactly as in
Exercise 2. Plot it against the posterior marginal to see the correspondence:

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
prec_marg <- as.data.frame(
  m_partial$marginals.hyperpar$`Precision for provincia_id`
)

ggplot(prec_marg, aes(x = x, y = y)) +
  geom_line(colour = "darkcyan", linewidth = 1) +
  geom_point(data = configs,
             aes(x = precision, y = 0, size = weight),
             colour = "#c0392b", alpha = 0.8, inherit.aes = FALSE) +
  scale_size_continuous(range = c(1, 6)) +
  labs(x = "Precision for provincia_id", y = "Density",
       size = "Weight")
```

### From precision to standard deviation

`INLA` reports the hyperparameter as a **precision**, $\tau = 1/\sigma_p^2$:

```{r}
m_partial$summary.hyperpar
```

A precision is difficult to interpret, and the posterior mean of
$1/\sqrt{\tau}$ is *not* $1/\sqrt{\text{posterior mean of } \tau}$. Transform
the whole marginal, not the summary:

```{r}
sd_marg <- inla.tmarginal(
  function(x) 1 / sqrt(x),
  m_partial$marginals.hyperpar$`Precision for provincia_id`
)

inla.zmarginal(sd_marg)
inla.emarginal(function(x) x, sd_marg)   # posterior mean of sigma_p
inla.hpdmarginal(0.95, sd_marg)          # 95% HPD interval
```

```{r}
ggplot(as.data.frame(sd_marg), aes(x = x, y = y)) +
  geom_line(colour = "darkcyan", linewidth = 1) +
  labs(x = expression(sigma[p]), y = "Density",
)
```

### Random effects and shrinkage

```{r}
re <- m_partial$summary.random$provincia_id

ggplot(re, aes(x = ID, y = mean)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_pointrange(aes(ymin = `0.025quant`, ymax = `0.975quant`),
                  linewidth = 0.2, size = 0.2) +
  labs(x = "Province", y = expression(theta[j]))
```

Now compare the unpooled and partially pooled estimates, province by province.
Provinces with small expected counts are pulled hardest, with the amount of
shrinkage estimated from the data via $\sigma_p$.

```{r}
shrinkage <- tibble(
  provincia_id = re$ID,
  unpooled     = m_unpooled$summary.fixed$mean,
  partial      = m_partial$summary.fixed$mean[1] + re$mean,
  expected     = data |> group_by(provincia_id) |> summarise(E = sum(expected)) |> pull(E)
)

ggplot(shrinkage, aes(x = unpooled, y = partial, size = expected)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  geom_point(alpha = 0.6, colour = "darkcyan") +
  scale_size_continuous(name = "Expected deaths") +
  labs(x = expression("Unpooled " * alpha[j]),
       y = expression("Partially pooled " * alpha + theta[j]),
)
```

### Map the partially pooled SMR

```{r}
SMR_partial_pooling <- exp(m_partial$summary.fixed$`0.5quant`[1] + re$`0.5quant`)

p_partial <- shp_italy |>
  mutate(SMR = SMR_partial_pooling) |>
  ggplot(aes(fill = SMR)) +
  geom_sf(colour = "white") +
  scale_fill_continuous_sequential(palette = "Reds") +
  theme_void()

p_partial
```

## Comparing each of the models

```{r}
p_full + labs(caption = "Full pooling model")
p_no + labs(caption = "No pooling model")
p_partial + labs(caption = "Partial pooling model")
```

There is clearly more smoothing in the partial pooling model.

```{r}
tibble(
  model = c("Full pooling", "No pooling", "Partial pooling"),
  dic   = map_dbl(list(m_pooled, m_unpooled, m_partial), ~ .x$dic$dic),
  waic  = map_dbl(list(m_pooled, m_unpooled, m_partial), ~ .x$waic$waic),
  p_eff = map_dbl(list(m_pooled, m_unpooled, m_partial), ~ .x$waic$p.eff)
) |> kable(digits = 1)
```

`p.eff` is the effective number of parameters. The partial pooling model has
107 province effects but far fewer *effective* parameters, because they are
tied together through $\sigma_p$.

`INLA` also provides leave-one-out quantities via CPO at little extra cost,
which would require refitting the model $n$ times with MCMC:

```{r}
# Leave-one-out log score (lower is better)
-mean(log(m_partial$cpo$cpo), na.rm = TRUE)

# Check the CPO computation is trustworthy (should be 0 failures)
sum(m_partial$cpo$failure > 0)
```

# Part 3: the approximation choices

The lecture described two separate dials. They control different stages and
are easy to confuse; the next two fits move them separately.

## `int.strategy`: how many combinations

This governs **stage 1**, where the hyperparameter space gets evaluated. It
is the choice you made by hand when you set `K` in Part 1.

```{r}
fit_int <- function(strategy) {
  system.time(
    m <- inla(
      deaths ~ 1 + f(provincia_id, model = "iid", hyper = pc_prec),
      family = "poisson",
      E      = expected,
      data   = data,
      control.fixed   = list(mean.intercept = 0, prec.intercept = 0.04),
      control.compute = list(config = TRUE),
      control.inla    = list(int.strategy = strategy)
    )
  ) -> tt
  list(fit = m, time = tt[["elapsed"]])
}

r_eb   <- fit_int("eb")
r_ccd  <- fit_int("ccd")
r_grid <- fit_int("grid")
```

```{r}
int_compare <- tibble(
  int.strategy = c("eb", "ccd", "grid"),
  n_points     = map_dbl(list(r_eb, r_ccd, r_grid),
                         ~ .x$fit$misc$configs$nconfig),
  seconds      = map_dbl(list(r_eb, r_ccd, r_grid), ~ .x$time),
  alpha_sd     = map_dbl(list(r_eb, r_ccd, r_grid),
                         ~ .x$fit$summary.fixed$sd[1]),
  ci_width     = map_dbl(list(r_eb, r_ccd, r_grid),
                         ~ diff(unlist(.x$fit$summary.random$provincia_id[1,
                             c("0.025quant", "0.975quant")])))
)

int_compare |> kable(digits = 4)
```

Look at `ci_width`. `eb` uses one combination and its intervals come out
**narrowest**, because the uncertainty in $\sigma_p$ never reaches the answer.
This is the red curve from Part 1, in a real model.

::: aside
With only one hyperparameter the differences are small, there is not much
uncertainty to lose. Re-run this comparison after fitting the BYM2 model
below, which has two, and the gap widens.
:::

**Practical habit:** develop with `eb`, report with `ccd`.

## `strategy`: how each marginal is read out

This governs a different stage, extracting an individual element's marginal
from the joint Gaussian. This is where the skew you saw in the Gamma example
gets corrected.

```{r}
fit_strat <- function(strategy) {
  system.time(
    m <- inla(
      deaths ~ 1 + f(provincia_id, model = "iid", hyper = pc_prec),
      family = "poisson",
      E      = expected,
      data   = data,
      control.fixed   = list(mean.intercept = 0, prec.intercept = 0.04),
      control.inla    = list(strategy = strategy),
      control.compute = list(return.marginals.random = TRUE)
    )
  ) -> tt
  list(fit = m, time = tt[["elapsed"]])
}

s_gauss <- fit_strat("gaussian")
s_simpl <- fit_strat("simplified.laplace")
s_lap   <- fit_strat("laplace")
```

```{r}
tibble(
  strategy  = c("gaussian", "simplified.laplace", "laplace"),
  seconds   = map_dbl(list(s_gauss, s_simpl, s_lap), ~ .x$time),
  mean_kld  = map_dbl(list(s_gauss, s_simpl, s_lap),
                      ~ mean(.x$fit$summary.random$provincia_id$kld)),
  max_kld   = map_dbl(list(s_gauss, s_simpl, s_lap),
                      ~ max(.x$fit$summary.random$provincia_id$kld))
) |> kable(digits = 6)
```

The `kld` column measures how far each marginal is from the Gaussian
approximation. Under `gaussian` it is zero **by construction**, the strategy
does not correct anything, so it cannot report a discrepancy. Under the other
two it tells you how much correcting was needed.

Compare a single province's marginal across strategies:

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
# the province with the least data: use its provincia_id, not its row number
j <- data$provincia_id[which.min(data$expected)]

# marginals.random is only returned when asked for (see control.compute above)
stopifnot(length(s_gauss$fit$marginals.random$provincia_id) >= j)

get_marg <- function(res, label) {
  res$fit$marginals.random$provincia_id[[j]] |>
    as.data.frame() |>
    mutate(strategy = label)
}

marg_df <- bind_rows(
  get_marg(s_gauss, "gaussian"),
  get_marg(s_simpl, "simplified.laplace"),
  get_marg(s_lap,   "laplace")
)

ggplot(marg_df, aes(x = x, y = y, colour = strategy, linetype = strategy)) +
  geom_line(linewidth = 0.9) +
  labs(x = expression(theta[j]), y = "Density")
```

The three curves should nearly coincide here. They separate when data per
latent element are thin. The mechanism is the same as `shape = 2` in
Exercise 1.

::: {.callout-tip}
## Exercise 3

- Why is `mean_kld` exactly zero under `strategy = "gaussian"`? What would you
  need to compute in order to know whether that strategy was adequate?
- The `laplace` strategy is the most expensive. On this model the extra cost
  is small. What feature of this dataset makes that true, and what kind of
  outcome would change it?
:::

## Prior sensitivity on the variance component

The default `INLA` prior on random effect precisions is `loggamma(1, 5e-5)`.
Widely used, but more informative than expected for small variances. Three
priors:

- PC prior `param = c(1, 0.01)`: $P(\sigma_p > 1) = 0.01$ (what we used);
- PC prior `param = c(0.1, 0.01)`: much tighter, allowing little
  between-province variation;
- the `loggamma` default.

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| out-width: "100%"
fit_partial <- function(hyper) {
  inla(
    deaths ~ 1 + f(provincia_id, model = "iid", hyper = hyper),
    family = "poisson",
    E      = expected,
    data   = data,
    control.fixed   = list(mean.intercept = 0, prec.intercept = 0.04),
    control.compute = list(dic = TRUE, waic = TRUE)
  )
}

m_pc_wide  <- fit_partial(list(prec = list(prior = "pc.prec", param = c(1, 0.01))))
m_pc_tight <- fit_partial(list(prec = list(prior = "pc.prec", param = c(0.1, 0.01))))
m_lgamma   <- fit_partial(list(prec = list(prior = "loggamma", param = c(1, 5e-5))))

sd_of <- function(m) {
  inla.tmarginal(function(x) 1 / sqrt(x),
                 m$marginals.hyperpar$`Precision for provincia_id`)
}

sens <- bind_rows(
  as.data.frame(sd_of(m_pc_wide))  |> mutate(prior = "PC, U = 1"),
  as.data.frame(sd_of(m_pc_tight)) |> mutate(prior = "PC, U = 0.1"),
  as.data.frame(sd_of(m_lgamma))   |> mutate(prior = "loggamma default")
)



p_a <- ggplot(sens, aes(x = x, y = y, colour = prior)) +
  geom_line(linewidth = 0.9) +
  xlim(0, 0.6) +
  labs(x = expression(sigma[p]), y = "Density")

smr_of <- function(m) {
  exp(m$summary.fixed$`0.5quant`[1] + m$summary.random$provincia_id$`0.5quant`)
}

p_b <- tibble(
  wide  = smr_of(m_pc_wide),
  tight = smr_of(m_pc_tight),
  lg    = smr_of(m_lgamma)
) |>
  pivot_longer(everything(), names_to = "prior", values_to = "SMR") |>
  ggplot(aes(x = SMR, colour = prior)) +
  geom_density(linewidth = 0.9) +
  labs(fill = "SMR")

p_a + p_b
```

With 107 provinces and reasonably large expected counts, the data are
informative and the SMRs move little; the posterior of $\sigma_p$
itself will typically move more than the SMRs do. Priors on variance
components should always be stated in a paper and checked with a sensitivity
analysis; with `INLA` each refit takes seconds.

# Part 4: sparsity and the spatial extension

## Looking at $\boldsymbol{Q}$

The lecture argued that the latent field is a GMRF, and that its precision
matrix is almost entirely blank. Now we can see it.

Build the adjacency structure from the shapefile using `spdep`, and write it
in the format `INLA` expects:

```{r}
nb <- poly2nb(shp_italy)
nb2INLA(here("data", "italy", "italy.adj"), nb)
g <- inla.read.graph(here("data", "italy", "italy.adj"))

summary(g)
```

The graph *is* the pattern of blanks. Look at it directly:

```{r}
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
inla.spy(g)
```

And count:

```{r}
Q_struct <- inla.graph2matrix(g)
Np       <- g$n

tibble(
  quantity = c("elements in the field", "entries in Q", "filled entries",
               "percentage filled"),
  value    = c(Np, Np^2, Matrix::nnzero(Q_struct),
               100 * Matrix::nnzero(Q_struct) / Np^2)
) |> kable(digits = 2)
```

Roughly six entries per row and several hundred blanks, with 107 provinces
the saving is modest, but the cost of a dense factorisation grows as $m^3$
while a sparse one grows closer to $m^{3/2}$. At 8,000 areas, the difference
between those is the difference between seconds and hours.

::: aside
Check `g$n` matches the number of provinces, and look out for islands (areas
with no neighbours), Italy has a couple, which `INLA` handles but which you
should know about. `summary(nb)` will tell you.
:::

## Spatial extension: BYM2

The IID model shrinks every province towards the *national* mean, ignoring
the map. BYM2 adds spatially structured smoothing: each province is pulled
towards its **neighbours** as well as the national mean.

$$
\begin{split}
\log(\mu_i) &= \log(E_i) + \alpha + b_{j[i]} \\
\boldsymbol{b} &= \frac{1}{\sqrt{\tau_b}}\left(\sqrt{1-\phi}\,\boldsymbol{v} + \sqrt{\phi}\,\boldsymbol{u}_\star\right)
\end{split}
$$

where $\boldsymbol{v}$ is IID noise, $\boldsymbol{u}_\star$ is a scaled ICAR
(neighbours-only) field, and $\phi \in [0, 1]$ is the proportion of the
variance that is spatially structured. $\phi = 0$ recovers the IID model;
$\phi = 1$ is pure ICAR.

This is our first model with **two** hyperparameters, so stage 1 now scans a
two-dimensional space rather than a line.

`scale.model = TRUE` scales the ICAR component so the precision means the same
thing regardless of the graph, essential for the PC prior to be
interpretable.

```{r}
t_bym2 <- system.time(
  m_bym2 <- inla(
    deaths ~ 1 + f(provincia_id, model = "bym2", graph = g,
                   scale.model = TRUE,
                   hyper = list(
                     prec = list(prior = "pc.prec", param = c(1, 0.01)),
                     phi  = list(prior = "pc",      param = c(0.5, 0.5)))),
    family = "poisson",
    E      = expected,
    data   = data,
    control.fixed   = list(mean.intercept = 0, prec.intercept = 0.04),
    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
  )
)

t_bym2
m_bym2$summary.hyperpar
```

How many combinations did it take this time?

```{r}
tibble(
  model            = c("Partial pooling (iid)", "BYM2"),
  hyperparameters  = c(1, 2),
  n_configurations = c(m_partial$misc$configs$nconfig,
                       m_bym2$misc$configs$nconfig),
  seconds          = c(t_partial_inla[["elapsed"]], t_bym2[["elapsed"]])
) |> kable(digits = 2)
```

One extra hyperparameter, and the number of points `INLA` must evaluate jumps.
This is the geometric growth from the lecture, in miniature. It is the reason
the practical ceiling is on the *number of hyperparameters*, not the size of
the latent field.

See the two-dimensional scan directly:

```{r}
#| fig-width: 6
#| fig-height: 5
#| fig-align: center
th <- map_dfr(m_bym2$misc$configs$config, \(cc) {
  tibble(log_prec = cc$theta[1], logit_phi = cc$theta[2],
         lp = cc$log.posterior)
}) |>
  mutate(weight = exp(lp - max(lp)), weight = weight / sum(weight))

ggplot(th, aes(x = log_prec, y = logit_phi, size = weight)) +
  geom_point(colour = "#c0392b", alpha = 0.7) +
  labs(x = "log precision", y = "logit phi",
       size = "Weight")
```

That scatter is the `ccd` design from the lecture: the mode, plus axial and
factorial points placed using the curvature.

What does the posterior of $\phi$ tell us? Values near 1 say most of the
between-province variation is spatially structured; values near 0 say the map
adds little beyond IID effects.

```{r}
phi_marg <- m_bym2$marginals.hyperpar$`Phi for provincia_id`

inla.zmarginal(phi_marg)

ggplot(as.data.frame(phi_marg), aes(x = x, y = y)) +
  geom_line(colour = "darkcyan", linewidth = 1) +
  labs(x = expression(phi), y = "Density",
)
```

Map the BYM2-smoothed SMRs next to the IID version. In the `bym2` model,
`summary.random` has $2 N_p$ rows: the first $N_p$ are the **combined** effect
$b_j$ (what we want for the map), the second $N_p$ are the structured
component alone.

```{r}
re_bym2 <- m_bym2$summary.random$provincia_id[1:Np, ]

SMR_bym2 <- exp(m_bym2$summary.fixed$`0.5quant`[1] + re_bym2$`0.5quant`)

p_bym2 <- shp_italy |>
  mutate(SMR = SMR_bym2) |>
  ggplot(aes(fill = SMR)) +
  geom_sf(colour = "white") +
  scale_fill_continuous_sequential(palette = "Reds") +
  theme_void()

(p_partial + labs(caption = "IID (partial pooling)")) +
  (p_bym2 + labs(caption = "BYM2 (spatially structured)"))
```

```{r}
tibble(
  model = c("Partial pooling (iid)", "BYM2"),
  dic   = map_dbl(list(m_partial, m_bym2), ~ .x$dic$dic),
  waic  = map_dbl(list(m_partial, m_bym2), ~ .x$waic$waic)
) |> kable(digits = 1)
```

::: {.callout-tip}
## Exercise 4

Re-run the `int.strategy` comparison from Part 3 on the BYM2 model, which has
two hyperparameters rather than one.

- How much wider are the `ccd` intervals than the `eb` ones now?
- Why should the gap grow with the number of hyperparameters?
:::

<!-- ## Adding time: a third hyperparameter -->

<!-- Everything so far used a single month. Take the whole of 2018 and add a smooth -->
<!-- temporal trend, and we arrive at the lecture's running example: a Poisson -->
<!-- likelihood, a spatial term and a temporal term. -->

<!-- ```{r} -->
<!-- data_year <- read_rds(here("data", "italy", "italy_mortality.rds")) |> -->
<!--   mutate(provincia_id = _ |> group_by(SIGLA) |> group_indices()) |> -->
<!--   filter(year == 2018) |> -->
<!--   arrange(SIGLA, month) -->

<!-- t_st <- system.time( -->
<!--   m_st <- inla( -->
<!--     deaths ~ 1 + -->
<!--       f(provincia_id, model = "bym2", graph = g, scale.model = TRUE, -->
<!--         hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01)), -->
<!--                      phi  = list(prior = "pc",      param = c(0.5, 0.5)))) + -->
<!--       f(month, model = "rw1", scale.model = TRUE, constr = TRUE, -->
<!--         hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01)))), -->
<!--     family = "poisson", -->
<!--     E      = expected, -->
<!--     data   = data_year, -->
<!--     control.fixed   = list(mean.intercept = 0, prec.intercept = 0.04), -->
<!--     control.compute = list(dic = TRUE, waic = TRUE, config = TRUE) -->
<!--   ) -->
<!-- ) -->

<!-- t_st -->
<!-- m_st$summary.hyperpar -->
<!-- ``` -->

<!-- ```{r} -->
<!-- tibble( -->
<!--   model            = c("Full pooling", "Partial pooling", "BYM2", "BYM2 + RW1"), -->
<!--   hyperparameters  = c(0, 1, 2, 3), -->
<!--   n_configurations = c(1, m_partial$misc$configs$nconfig, -->
<!--                        m_bym2$misc$configs$nconfig, -->
<!--                        m_st$misc$configs$nconfig), -->
<!--   seconds          = c(t_full_inla[["elapsed"]], t_partial_inla[["elapsed"]], -->
<!--                        t_bym2[["elapsed"]], t_st[["elapsed"]]) -->
<!-- ) |> kable(digits = 2) -->
<!-- ``` -->

<!-- The latent field grew twelvefold when we added the other eleven months. The -->
<!-- runtime is driven by the third column, not the size of the field. -->

<!-- The temporal trend itself: -->

<!-- ```{r} -->
<!-- ggplot(m_st$summary.random$month, -->
<!--        aes(x = ID, y = mean)) + -->
<!--   geom_hline(yintercept = 0, linetype = 2) + -->
<!--   geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), -->
<!--               alpha = 0.2, fill = "darkcyan") + -->
<!--   geom_line(colour = "darkcyan", linewidth = 1) + -->
<!--   scale_x_continuous(breaks = 1:12) + -->
<!--   labs(x = "Month of 2018", y = "RW1 effect", -->
<!--        title = "Smooth seasonal trend in mortality") -->
<!-- ``` -->

<!-- # Part 5: comparing INLA and NIMBLE -->

<!-- Finally, the two approaches side by side. Below we refit the partial pooling -->
<!-- model with `nimbleMCMC()` exactly as in the hierarchical modelling lab. -->

<!-- ::: aside -->
<!-- The `NIMBLE` fit takes a couple of minutes, most of it model compilation. Feel -->
<!-- free to skip re-running it and use your saved output from the previous lab. -->
<!-- ::: -->

<!-- ```{r} -->
<!-- #| cache: true -->
<!-- constants <- list( -->
<!--   N = nrow(data), -->
<!--   Np = max(data$provincia_id), -->
<!--   province = data$provincia_id -->
<!-- ) -->

<!-- nimble_data <- list( -->
<!--   y = as.integer(data$deaths), -->
<!--   E = data$expected -->
<!-- ) -->

<!-- inits <- list( -->
<!--   list(alpha = 0,   theta = rep(0, constants$Np), sigma_p = 1), -->
<!--   list(alpha = 0.1, theta = rep(0, constants$Np), sigma_p = 1) -->
<!-- ) -->

<!-- partial_pooling_model <- nimbleCode({ -->
<!--   alpha ~ dnorm(0, 5) -->
<!--   sigma_p ~ T(dnorm(0, 1), 0, Inf) -->

<!--   for (j in 1:Np) { -->
<!--     theta[j] ~ dnorm(0, sd = sigma_p) -->
<!--   } -->

<!--   for (i in 1:N) { -->
<!--     y[i] ~ dpois(mu[i]) -->
<!--     log(mu[i]) <- log(E[i]) + alpha + theta[province[i]] -->
<!--   } -->
<!-- }) -->

<!-- t_partial_nimble <- system.time( -->
<!--   partial_pooling_samples <- nimbleMCMC( -->
<!--     code = partial_pooling_model, -->
<!--     data = nimble_data, -->
<!--     constants = constants, -->
<!--     inits = inits, -->
<!--     monitors = c("alpha", "sigma_p", "theta"), -->
<!--     niter = 2000, -->
<!--     nburnin = 1000, -->
<!--     thin = 1, -->
<!--     nchains = 2, -->
<!--     samplesAsCodaMCMC = TRUE, -->
<!--     progressBar = FALSE, -->
<!--     WAIC = TRUE -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- tibble( -->
<!--   engine  = c("NIMBLE (MCMC)", "INLA"), -->
<!--   elapsed = c(t_partial_nimble["elapsed"], t_partial_inla["elapsed"]) -->
<!-- ) |> kable(digits = 2) -->
<!-- ``` -->

<!-- Note what each engine was doing. `NIMBLE` sampled 108 latent quantities *and* -->
<!-- the hyperparameter jointly, 2,000 times, twice. `INLA` scanned a handful of -->
<!-- values of $\sigma_p$ and solved the latent field once at each. The second -->
<!-- column is the whole argument of the lecture. -->

<!-- And do they agree? -->

<!-- ```{r} -->
<!-- nimble_theta <- summarise_draws( -->
<!--   partial_pooling_samples$samples, -->
<!--   default_summary_measures() -->
<!-- ) |> -->
<!--   filter(str_detect(variable, "theta")) |> -->
<!--   arrange(as.integer(str_extract(variable, "\\d+"))) -->

<!-- agreement <- tibble( -->
<!--   inla   = m_partial$summary.random$provincia_id$mean, -->
<!--   nimble = nimble_theta$mean -->
<!-- ) -->

<!-- ggplot(agreement, aes(x = nimble, y = inla)) + -->
<!--   geom_abline(slope = 1, intercept = 0, linetype = 2) + -->
<!--   geom_point(alpha = 0.6, size = 0.8, colour = "darkcyan") + -->
<!--   labs(x = expression("NIMBLE posterior mean of " * theta[j]), -->
<!--        y = expression("INLA posterior mean of " * theta[j])) -->
<!-- ``` -->

<!-- The points should sit close to the line; any scatter is mostly Monte Carlo -->
<!-- noise in the (short) `NIMBLE` run, plus the difference between the half-normal -->
<!-- and PC priors on $\sigma_p$. MCMC error shrinks as the chains run longer, -->
<!-- whereas the `INLA` approximation error is fixed but, for models like this, -->
<!-- very small. -->

<!-- ::: {.callout-tip} -->
<!-- ## Exercise 5 -->

<!-- The `NIMBLE` model above has one hyperparameter (`sigma_p`) and 108 latent -->
<!-- quantities, and samples them together. Suppose you wanted 8,000 areas over 20 -->
<!-- years with a BYM2 field and an RW1 trend. -->

<!-- - Which engine's cost grows fastest, and with respect to what? -->
<!-- - Which of the lecture's two conditions would you check first before reaching -->
<!--   for `INLA`? -->
<!-- ::: -->

## Closing remarks

In this lab we built both halves of INLA by hand: a Laplace approximation
from a mode and a curvature, and a weighted sum over a grid of hyperparameter
values. `R-INLA` then did the same on real data, at a scale where doing it by
hand is not possible.

The two conditions from the lecture held throughout. Every model here had a
Gaussian latent field with local links, and between zero and three
hyperparameters. For models with many variance components, or a latent field
that is not Gaussian, MCMC in `NIMBLE` remains the reference approach.

A common workflow is to develop and compare models in `INLA`, using the speed
for prior sensitivity and cross-validation, then verify the final model
with MCMC where the extra flexibility or exactness is needed.
