---
title: "Priors in Bayesian Environmental Health Models"
subtitle: "SHARP Bayesian Modeling for Environmental Health Workshop"
author: "Garyfallos Konstantinoudis"
date: "August 5 2026"
format: html 

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(nimble)
library(coda)
library(dlnm)
library(patchwork)
library(posterior)
library(bayesplot)
library(here)
library(hrbrthemes)

theme_set(theme_ipsum())
knitr::opts_chunk$set(fig.align = "center", warning = FALSE, message = FALSE)
set.seed(2)
```

## Goal of this computing lab session

The goal of this lab is to explore how different prior choices affect posterior inference in simple Bayesian environmental health models using NIMBLE.

## What's going to happen in this lab session?

During this lab session, we will:

- Fit a simple Bayesian Poisson regression model for ozone and mortality in Chicago;
- Compare posterior inference under different priors for the ozone effect;
- Fit a Bayesian linear-threshold model for summer temperature and mortality;
- Compare how different priors on the threshold affect posterior inference for the threshold and the slope;
- Use basic MCMC diagnostics to assess convergence and sampling quality.

## Introduction

In Bayesian statistics, priors are a central part of the model. In practice, priors can affect inference in different ways depending on:

- how informative the data are,
- how strongly the prior is specified,
- and whether the parameter is a regression coefficient or a structural parameter.

In this session, we use the `chicagoNMMAPS` dataset from the `dlnm` package to study two examples:

- a simple ozone-mortality model, where we vary the prior on the regression coefficient;
- a summer temperature threshold model, where we vary the prior on the threshold parameter.

These examples show how Bayesian updating works in practice and how prior sensitivity can be assessed with simple models.

## Example 1: ozone and all-cause mortality

We begin with a simple Poisson regression model for daily mortality and ozone.

Let $Y_t$ denote daily all-cause mortality and let $x_t$ denote ozone concentration. We fit the model:

$$
Y_t \sim \text{Poisson}(\mu_t)
$$

$$
\log(\mu_t) = \alpha + \beta x_t
$$

where $\beta$ is the log-relative risk associated with ozone.

Our goal is to compare posterior inference under different priors for $\beta$.

### Load and prepare the Chicago data

```{r}
data("chicagoNMMAPS")

dat <- chicagoNMMAPS %>%
  select(date, death, o3) %>%
  na.omit()

dat$date <- as.Date(dat$date)
dat$x_o3 <- dat$o3 / 10

y <- dat$death
x <- dat$x_o3
n <- length(y)

glimpse(dat)
summary(dat)
```

### Define the NIMBLE model

```{r}
airpoll_code <- nimbleCode({
  for (t in 1:n) {
    y[t] ~ dpois(mu[t])
    log(mu[t]) <- alpha + beta * x[t]
  }

  alpha ~ dnorm(0, sd = 100)
  beta  ~ dnorm(beta_mean, sd = beta_sd)
})
```

### Fit the model under different priors

We compare:

- a diffuse prior,
- a weakly informative prior,
- an informative prior.

```{r}
fit_nimble_model <- function(beta_mean, beta_sd,
                             niter = 12000, nburnin = 2000, nthin = 10,
                             nchains = 2, seed = 2) {

  constants <- list(
    n = n,
    x = x,
    beta_mean = beta_mean,
    beta_sd = beta_sd
  )

  data_list <- list(y = y)

  inits <- list(
    list(alpha = log(mean(y)), beta = 0),
    list(alpha = log(mean(y)) + 0.1, beta = 0.02)
  )

  model <- nimbleModel(
    code = airpoll_code,
    constants = constants,
    data = data_list,
    inits = inits[[1]]
  )

  cmodel <- compileNimble(model)

  conf <- configureMCMC(model, monitors = c("alpha", "beta"))
  mcmc <- buildMCMC(conf)
  cmcmc <- compileNimble(mcmc, project = model)

  samples <- runMCMC(
    cmcmc,
    niter = niter,
    nburnin = nburnin,
    thin = nthin,
    nchains = nchains,
    inits = inits,
    setSeed = c(seed, seed + 1),
    samplesAsCodaMCMC = TRUE,
    summary = TRUE
  )

  return(samples)
}
```

```{r eval=FALSE}
fit_diffuse <- fit_nimble_model(beta_mean = 0,    beta_sd = 0.5,   seed = 11)
fit_weak    <- fit_nimble_model(beta_mean = 0,    beta_sd = 0.1,   seed = 12)
fit_info    <- fit_nimble_model(beta_mean = 0.01, beta_sd = 0.005, seed = 13)

saveRDS(fit_diffuse, here::here('labs','priors','model_fits',"fit_diffuse.rds"))
saveRDS(fit_weak, here::here('labs','priors','model_fits',"fit_weak.rds"))
saveRDS(fit_info, here::here('labs','priors','model_fits',"fit_info.rds"))
```

```{r echo = FALSE}
fit_diffuse <- readRDS(here::here('labs','priors','model_fits',"fit_diffuse.rds"))
fit_weak    <- readRDS(here::here('labs','priors','model_fits',"fit_weak.rds"))
fit_info    <- readRDS(here::here('labs','priors','model_fits',"fit_info.rds"))
```

### Posterior summaries

### Diagnostics

We begin by checking convergence and mixing for $\beta$.

```{r}
summarise_draws(fit_diffuse$samples, default_summary_measures())
summarise_draws(fit_diffuse$samples, default_convergence_measures())
```

```{r}
mcmc_trace(fit_diffuse$samples, pars = "beta")
mcmc_trace(fit_weak$samples, pars = "beta")
mcmc_trace(fit_info$samples, pars = "beta")
```


### Extract posterior samples

```{r}
beta_diffuse <- as.matrix(fit_diffuse$samples)[, "beta"]
beta_weak    <- as.matrix(fit_weak$samples)[, "beta"]
beta_info    <- as.matrix(fit_info$samples)[, "beta"]
```

### Compare prior and posterior distributions

```{r}
setxlim <- quantile(c(beta_diffuse, beta_weak, beta_info), c(0.01, 0.99))
setxlim <- c(setxlim[1] - 0.01, setxlim[2] + 0.01)

make_density_df <- function(samples, mean_prior, sd_prior, label,
                            xlim = setxlim, n_grid = 500) {

  grid <- seq(xlim[1], xlim[2], length.out = n_grid)

  prior_df <- data.frame(
    beta = grid,
    density = dnorm(grid, mean = mean_prior, sd = sd_prior),
    distribution = "Prior",
    prior_type = label
  )

  post_dens <- density(samples, from = xlim[1], to = xlim[2], n = n_grid)

  post_df <- data.frame(
    beta = post_dens$x,
    density = post_dens$y,
    distribution = "Posterior",
    prior_type = label
  )

  bind_rows(prior_df, post_df)
}

plot_df_o3 <- bind_rows(
  make_density_df(beta_diffuse, mean_prior = 0,    sd_prior = 0.5,   label = "Diffuse"),
  make_density_df(beta_weak,    mean_prior = 0,    sd_prior = 0.1,   label = "Weakly informative"),
  make_density_df(beta_info,    mean_prior = 0.01, sd_prior = 0.005, label = "Informative")
)
```

```{r}
ggplot(plot_df_o3, aes(x = beta, y = density, color = prior_type)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  facet_wrap(~ distribution, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Diffuse" = "#E69F00",
    "Weakly informative" = "#0072B2",
    "Informative" = "#D55E00"
  )) +
  labs(
    x = expression(beta),
    y = "Density",
    color = "Prior choice",
    title = expression("Priors and posteriors for " * beta),
    subtitle = "Chicago ozone-mortality example"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

### Interpretation

This comparison shows how the posterior for $\beta$ depends on the prior.

In many real datasets, the posterior distributions are more similar than the prior distributions, indicating that the data are informative. However, stronger priors can still shift the posterior, especially when they encode substantial prior information.

## Example 2: summer temperature threshold model

We now consider a second example where the parameter of interest is not just a regression coefficient, but a threshold.

We restrict the Chicago data to summer months and fit a linear-threshold model for temperature:

$$
Y_t \sim \text{Poisson}(\mu_t)
$$

$$
\log(\mu_t) = \alpha + \beta (T_t - \psi)_+
$$

where:

- $T_t$ is summer temperature,
- $\psi$ is the threshold,
- $\beta$ is the slope above the threshold.

Our goal is to see how the prior on $\psi$ affects:

- the posterior of the threshold,
- the posterior of the slope,
- and the fitted threshold-response function.

### Restrict to summer data

```{r}
dat2 <- chicagoNMMAPS
dat2$date <- as.Date(dat2$date)

dat2$month <- as.integer(format(dat2$date, "%m"))
dat_summer <- dat2 %>%
  filter(month %in% c(6, 7, 8)) %>%
  select(date, death, temp) %>%
  na.omit()

y2   <- dat_summer$death
temp <- dat_summer$temp
n2   <- length(y2)

glimpse(dat_summer)
summary(dat_summer)
range(temp)
```

### Define the threshold model in NIMBLE

```{r}
threshold_code_unif <- nimbleCode({
  for (t in 1:n2) {
    y2[t] ~ dpois(mu[t])
    log(mu[t]) <- alpha + beta * (temp[t] - psi) * step(temp[t] - psi)
  }

  alpha ~ dnorm(0, sd = 100)
  beta  ~ dnorm(0, sd = 0.1)
  psi   ~ dunif(psi_lower, psi_upper)
})
```

### Fit the model under different threshold priors

We compare three uniform priors for the threshold:

- broad prior,
- lower prior,
- higher prior.

```{r}
fit_threshold_uniform <- function(psi_lower, psi_upper,
                                  niter = 15000,
                                  nburnin = 3000,
                                  nthin = 10,
                                  nchains = 2,
                                  seed = 2) {

  constants <- list(
    n2 = n2,
    temp = temp,
    psi_lower = psi_lower,
    psi_upper = psi_upper
  )

  data_list <- list(
    y2 = y2
  )

  inits <- list(
    list(alpha = log(mean(y2)), beta = 0.01, psi = mean(c(psi_lower, psi_upper))),
    list(alpha = log(mean(y2)) + 0.1, beta = 0.02, psi = mean(c(psi_lower, psi_upper)) + 0.5)
  )

  model <- nimbleModel(
    code = threshold_code_unif,
    constants = constants,
    data = data_list,
    inits = inits[[1]]
  )

  cmodel <- compileNimble(model)

  conf <- configureMCMC(model, monitors = c("alpha", "beta", "psi"))
  mcmc <- buildMCMC(conf)
  cmcmc <- compileNimble(mcmc, project = model)

  samples <- runMCMC(
    cmcmc,
    niter = niter,
    nburnin = nburnin,
    thin = nthin,
    nchains = nchains,
    inits = inits,
    setSeed = c(seed, seed + 1),
    samplesAsCodaMCMC = TRUE,
    summary = TRUE
  )

  return(samples)
}
```

```{r eval=FALSE}
fit_psi_broad <- fit_threshold_uniform(psi_lower = 20, psi_upper = 35, seed = 21)
fit_psi_low   <- fit_threshold_uniform(psi_lower = 24, psi_upper = 28, seed = 22)
fit_psi_high  <- fit_threshold_uniform(psi_lower = 29, psi_upper = 31, seed = 23)

saveRDS(fit_psi_broad, here::here('labs','priors','model_fits',"fit_psi_broad.rds"))
saveRDS(fit_psi_low,   here::here('labs','priors','model_fits',"fit_psi_low.rds"))
saveRDS(fit_psi_high,  here::here('labs','priors','model_fits',"fit_psi_high.rds"))
```

```{r echo = FALSE}
fit_psi_broad <- readRDS(here::here('labs','priors','model_fits',"fit_psi_broad.rds"))
fit_psi_low   <- readRDS(here::here('labs','priors','model_fits',"fit_psi_low.rds"))
fit_psi_high  <- readRDS(here::here('labs','priors','model_fits',"fit_psi_high.rds"))
```

### Diagnostics

Threshold models can mix more slowly than simple regression models, so it is useful to inspect the chains carefully.


```{r}

mcmc_trace(fit_psi_broad$samples, pars = c("beta", "psi"))
mcmc_trace(fit_psi_low$samples, pars = c("beta", "psi"))
mcmc_trace(fit_psi_high$samples, pars = c("beta", "psi"))

```

### Extract posterior samples

```{r}
samp_broad <- as.matrix(fit_psi_broad$samples)
samp_low   <- as.matrix(fit_psi_low$samples)
samp_high  <- as.matrix(fit_psi_high$samples)

psi_broad  <- samp_broad[, "psi"]
psi_low    <- samp_low[, "psi"]
psi_high   <- samp_high[, "psi"]

beta_broad <- samp_broad[, "beta"]
beta_low   <- samp_low[, "beta"]
beta_high  <- samp_high[, "beta"]
```

### Prepare summaries for plotting

```{r}
make_psi_df <- function(samples, lower, upper, label,
                        xlim = range(temp), n_grid = 500) {

  grid <- seq(xlim[1], xlim[2], length.out = n_grid)

  prior_df <- data.frame(
    value = grid,
    density = dunif(grid, min = lower, max = upper),
    distribution = "Prior for threshold",
    prior_type = label,
    parameter = "psi"
  )

  post_dens <- density(samples, from = xlim[1], to = xlim[2], n = n_grid)

  post_df <- data.frame(
    value = post_dens$x,
    density = post_dens$y,
    distribution = "Posterior for threshold",
    prior_type = label,
    parameter = "psi"
  )

  bind_rows(prior_df, post_df)
}

make_beta_df <- function(samples, label, xlim = NULL, n_grid = 500) {

  if (is.null(xlim)) {
    xlim <- range(samples)
  }

  post_dens <- density(samples, from = xlim[1], to = xlim[2], n = n_grid)

  data.frame(
    value = post_dens$x,
    density = post_dens$y,
    distribution = "Posterior for slope above threshold",
    prior_type = label,
    parameter = "beta"
  )
}
```

```{r}
psi_df <- bind_rows(
  make_psi_df(psi_broad, 20, 35, "Broad prior"),
  make_psi_df(psi_low,   24, 28, "Lower prior"),
  make_psi_df(psi_high,  29, 31, "Higher prior")
)

beta_xlim <- range(c(beta_broad, beta_low, beta_high))

beta_df <- bind_rows(
  make_beta_df(beta_broad, "Broad prior", beta_xlim),
  make_beta_df(beta_low,   "Lower prior", beta_xlim),
  make_beta_df(beta_high,  "Higher prior", beta_xlim)
)
```

### Plot the threshold prior sensitivity analysis

```{r}
mean_psi_broad  <- mean(psi_broad)
mean_psi_low    <- mean(psi_low)
mean_psi_high   <- mean(psi_high)

mean_beta_broad <- mean(beta_broad)
mean_beta_low   <- mean(beta_low)
mean_beta_high  <- mean(beta_high)

temp_grid <- seq(min(temp), max(temp), length.out = 200)

curve_df <- bind_rows(
  data.frame(
    temp = temp_grid,
    value = exp(mean_beta_broad * pmax(temp_grid - mean_psi_broad, 0)),
    prior_type = "Broad prior"
  ),
  data.frame(
    temp = temp_grid,
    value = exp(mean_beta_low * pmax(temp_grid - mean_psi_low, 0)),
    prior_type = "Lower prior"
  ),
  data.frame(
    temp = temp_grid,
    value = exp(mean_beta_high * pmax(temp_grid - mean_psi_high, 0)),
    prior_type = "Higher prior"
  )
)

psi_prior_df <- psi_df %>%
  filter(distribution == "Prior for threshold") %>%
  transmute(x = value, density = density, prior_type = prior_type)

psi_post_df <- psi_df %>%
  filter(distribution == "Posterior for threshold") %>%
  transmute(x = value, density = density, prior_type = prior_type)

beta_plot_df <- beta_df %>%
  transmute(x = value, density = density, prior_type = prior_type)
```

```{r height = 15, width = 13}
p1 <- ggplot(psi_prior_df, aes(x = x, y = density, color = prior_type)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  scale_color_manual(values = c(
    "Broad prior"  = "#0072B2",
    "Lower prior"  = "#D55E00",
    "Higher prior" = "#009E73"
  )) +
  labs(title = "Prior for threshold", x = expression(psi), y = "Density") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p2 <- ggplot(psi_post_df, aes(x = x, y = density, color = prior_type)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  scale_color_manual(values = c(
    "Broad prior"  = "#0072B2",
    "Lower prior"  = "#D55E00",
    "Higher prior" = "#009E73"
  )) +
  labs(title = "Posterior for threshold", x = expression(psi), y = "Density") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p3 <- ggplot(beta_plot_df, aes(x = x, y = density, color = prior_type)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  scale_color_manual(values = c(
    "Broad prior"  = "#0072B2",
    "Lower prior"  = "#D55E00",
    "Higher prior" = "#009E73"
  )) +
  labs(title = "Posterior for slope", x = expression(beta), y = "Density") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p4 <- ggplot(curve_df, aes(x = temp, y = value, color = prior_type)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  scale_color_manual(values = c(
    "Broad prior"  = "#0072B2",
    "Lower prior"  = "#D55E00",
    "Higher prior" = "#009E73"
  )) +
  labs(
    title = "Estimated threshold-response",
    x = "Temperature",
    y = "Relative risk"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

((p1 | p2) / (p3 | p4)) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### Interpretation

This example shows that priors can matter not only for regression coefficients, but also for structural parameters such as thresholds.

Changing the prior on $\psi$ can affect:

- the posterior location of the threshold,
- the posterior of the slope above threshold,
- and the implied temperature-mortality relationship.

This is a useful reminder that prior sensitivity should be checked not only for coefficients, but also for model structure.

## Closing remarks

In this lab session, we explored two simple Bayesian models using the Chicago mortality dataset.

First, we examined how different priors for $\beta$ affect posterior inference in a Poisson regression model.

Second, we studied a summer temperature threshold model and showed how different priors for $\psi$ affect both the posterior threshold and the estimated slope above it.

Together, these examples illustrate an important practical point:

- prior choice should be explicit,
- priors should be chosen on an interpretable scale,
- and sensitivity analysis is an essential part of Bayesian workflow.