---
title: "Advanced Environmental Modelling"
subtitle: "SHARP Bayesian Modeling for Environmental Health Workshop"
author: "Garyfallos Konstantinoudis, Ellen Steward"
date: "August 06 2026"
format: html
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
library(tidyverse)
library(INLA)
library(sf)
library(spdep)
library(dlnm)
library(lubridate)
library(colorspace)
library(patchwork)

extrafont::loadfonts()
theme_set(hrbrthemes::theme_ipsum())
knitr::opts_chunk$set(fig.align = "center")

set.seed(2)
```

## Goal of this computing lab session

This goal of this lab is to use `INLA` to fit advanced environmental models. We selected to use `INLA` as it these models require a lot of time to run in `NIMBLE`.

## What's going to happen in this lab session?

During this lab session, we will:

1. Explore data on Dengue cases in Sao Paolo
2. Define the neighborhood matrix in `R` (specific for `INLA`); and
3. Fit and interpret advanced environmental models in `INLA`

## Introduction

We will use the Dengue incidence during 2009-2024, in Brazil, at the micro-region level (645 areas), as downloaded from DATASUS.

For that analysis, we included 5,201,154 confirmed dengue cases in Sao Paolo. We retrieved mean, min and max monthly temperature, total monthly precipitation and mean monthly relative humidity from ERA5.Population was retrieved from the 2011 census. 

We will use Bayesian hierarchical models to quantify the effect of meteorology while adjusting for spatial, temporal and spatiotemporal autocorrelation.

## Read the data

Let's load in the data
```{r}
data_Brazil <- readRDS("data/dataBrazil.rds")
glimpse(data_Brazil)
summary(data_Brazil)
class(data_Brazil)
```

and the shapefile

```{r}
shp <- read_sf("data/micro_sp.shp")
plot(shp$geometry)
```

## INLA models

WRITE THE MODEL!!

To include random effects in INLA, you need to create IDs. 
```{r}
# seasonality
data_Brazil$id_seasonal <- data_Brazil$month
# long-term trends
data_Brazil$id_year <- data_Brazil$year - min(data_Brazil$year) + 1
# spatial effects
data_Brazil$id_space <- data_Brazil$code_muni %>% as.factor() %>% as.numeric()
```

We need to make sure that the spatial shp is alligned with the data structure. That means that the region with id_space 1 needs to be the 1st row of the shapefile, the regions with id_space 2, needs to be the 2nd row of the shapefile, etc. 

```{r}
shp$id_space <- shp$code_muni %>% as.factor() %>% as.numeric()
shp %>% arrange(id_space) -> shp

# and once this is done, we need to create the neighbor matrix
W.nb <- poly2nb(shp)
# and store it in a format for INLA
nb2INLA("W.adj", W.nb) 
```

Select the priors:
```{r}
# Priors
hyper.iid <- list(theta = list(prior="pc.prec", param=c(1,0.01)))
hyper.bym <- list(theta1 = list('PCprior', c(1,0.01)), theta2 = list('PCprior', c(0.5, 0.5)))
```

Define the formula
```{r}
mod1 = cases ~ 1 + 
  offset(log(population)) +  
  
  # long-term trends
  f(id_year, model = "rw1", hyper = hyper.iid, constr = TRUE) + 
  # seasonality
  f(id_seasonal, model = "rw2", hyper = hyper.iid, constr = TRUE) + 

  # spatial effect
  f(id_space, model = "bym2", graph='W.adj', hyper = hyper.bym, constr = TRUE) + 
  
  # meteorological linear terms
  rh + t2m_max + tp_sum
```

and the INLA call:
```{r, echo = TRUE, eval = FALSE}
inla(formula = as.formula(mod1),
            data=data_Brazil,
            family="poisson",
            verbose = TRUE, 
            
            # priors fixed effects
            control.fixed = list(
              
              mean = list(rh = 0, t2m_max = 0, tp_sum = 0), 
              prec = list(rh = 0.01, t2m_max = 0.01, tp_sum = 0.01),
              prec.intercept = 0.001, mean.intercept = 0),
            
            control.compute=list(dic=TRUE, cpo=TRUE, config = TRUE), 
            control.inla=list(strategy="simplified.laplace", int.strategy="eb"),
            control.mode=list(restart = T), 
            num.threads = 1
  ) -> mod1_res # 23 seconds

saveRDS(mod1_res, file = "data/mod1.rds")
```

```{r, echo = FALSE, eval = TRUE}
mod1_res <- readRDS("data/mod1.rds")
```

Extract the INLA output:
```{r warning=FALSE}
# fixed effects
mod1_res$summary.fixed
# seasonal terms

data.frame(
  x = mod1_res$summary.random$id_seasonal$ID, 
  RR = exp(mod1_res$summary.random$id_seasonal$`0.5quant`), 
  UL = exp(mod1_res$summary.random$id_seasonal$`0.975quant`), 
  LL = exp(mod1_res$summary.random$id_seasonal$`0.025quant`)
) %>% 
  ggplot() + 
  geom_line(aes(x=x, y=RR)) + 
  geom_ribbon(aes(x=x, ymin =LL, ymax=UL), alpha = 0.2)

# long-term trends
data.frame(
  x = mod1_res$summary.random$id_year$ID, 
  RR = exp(mod1_res$summary.random$id_year$`0.5quant`), 
  UL = exp(mod1_res$summary.random$id_year$`0.975quant`), 
  LL = exp(mod1_res$summary.random$id_year$`0.025quant`)
) %>% 
  ggplot() + 
  geom_line(aes(x=x, y=RR)) + 
  geom_ribbon(aes(x=x, ymin =LL, ymax=UL), alpha = 0.2)

# spatial effects
shp$RR <- mod1_res$summary.random$id_space$`0.5quant`[1:645]
shp %>% 
  ggplot() + 
  geom_sf(aes(fill=RR), col = NA) +
  scale_fill_viridis_c()
```

## Spatially varying temperature

In INLA you cannot use twice the same id. If needed you need to create a new column with a different name:
```{r}
data_Brazil$id_space2 <- data_Brazil$id_space
```


Define the formula
```{r}
mod2 = cases ~ 1 + 
  offset(log(population)) +  
  
  # long-term trends
  f(id_year, model = "rw1", hyper = hyper.iid, constr = TRUE) + 
  # seasonality
  f(id_seasonal, model = "rw2", hyper = hyper.iid, constr = TRUE) + 

  # spatial effect
  f(id_space, model = "bym2", graph='W.adj', hyper = hyper.bym, constr = TRUE) + 
  
  # meteorological linear terms
  rh + t2m_max + tp_sum +
  
  f(id_space2, t2m_max, model = "bym2", graph='W.adj', hyper = hyper.bym, constr = TRUE) 
```

If you have defined a linear effect of the temperature, you would need to add constraints for the SVC. The interpretation would be about spatial deviations of the overall linear term. If you do not include a linear effect, then the SVC is the total effect, including the average across the area. 

Run INLA:
```{r, echo = TRUE, eval = FALSE}
inla(formula = as.formula(mod2),
            data=data_Brazil,
            family="poisson",
            verbose = TRUE, 
            
            # priors fixed effects
            control.fixed = list(
              
              mean = list(rh = 0, t2m_max = 0, tp_sum = 0), 
              prec = list(rh = 0.01, t2m_max = 0.01, tp_sum = 0.01),
              prec.intercept = 0.001, mean.intercept = 0),
            
            control.compute=list(dic=TRUE, cpo=TRUE, config = TRUE), 
            control.inla=list(strategy="simplified.laplace", int.strategy="eb"),
            control.mode=list(restart = T), 
            num.threads = 1
  ) -> mod2_res # 52 seconds

saveRDS(mod2_res, file = "data/mod2.rds")
```

```{r, echo = FALSE, eval = TRUE}
mod2_res <- readRDS("data/mod2.rds")
```


Similar as before, all the random effects defined in the formula can be accessed in the `summary.random`, whereas the fixed effects in `summary.fixed`:

```{r warning=FALSE, fig.width=12}
# fixed effects
mod2_res$summary.fixed
# seasonal terms

# spatial effects
shp$RR <- mod2_res$summary.random$id_space2$`0.5quant`[1:645]
shp %>% 
  ggplot() + 
  geom_sf(aes(fill=RR), col = NA) +
  scale_fill_viridis_c() -> p1

# calculate exceedance probability: 
shp$exceed <- 1 - sapply(mod2_res$marginals.random$id_space2, FUN = inla.pmarginal, q = 0)[1:645]

shp %>% 
  ggplot() + 
  geom_sf(aes(fill=exceed), col = NA) +
  scale_fill_viridis_c() -> p2


p1|p2
```
Your turn!

Following the previous steps: 

a. Modify the formula to add a spatially varying term for the rainfall. (HINT: don't forget to create a new ID)
```{r echo = FALSE, eval = FALSE, warning=FALSE}

```


b. Run `INLA` with the updated formula.
```{r echo = FALSE, eval = FALSE, warning=FALSE}

```


c. Extract the spatial varying rainfall and plot it together with the exceedance probability. What do you observe?
```{r echo = FALSE, eval = FALSE, warning=FALSE, fig.width=12}

```


## DLNM with INLA

In this subsection, we will fit a DLNM to allow for the effect of temperature to be non-linear and lagged. 

First we need to define the basis function: 
```{r}
n.lags <- 12 # 12 months lags

# non-linear lag effect
blag.args <- list(
  fun = "bs",
  df = 4, #lags 
  degree = 2
)

blag.args$knots <- do.call(
  "equalknots", c(list(x = 0:n.lags), blag.args)) 
blag.args$knots 
  
## temperature basis functions argument list
btemp.args <- list(
  fun = "bs",
  df = 4,
  degree = 2)
btemp.args$knots <- quantile(data_Brazil$t2m_mean, 
                             probs = c(0.20, 0.80), 
                             na.rm = TRUE) 
btemp.args$knots
```

Calculate the crossbasis by region.
```{r}

# define the chronological sequence to calculate lags
data_Brazil$c_lags <- paste(data_Brazil$year, sprintf("%02d", data_Brazil$month), sep = "-") %>% 
  as.factor() %>% 
  as.numeric()

btemp.locs <- lapply(
    split(x = data_Brazil[c("c_lags", "code_muni", "t2m_mean")], 
          f = data_Brazil$code_muni), function(d) {
            data.frame(
              d,
              crossbasis(d$t2m_mean,
                         lag = n.lags,
                         argvar = btemp.args,
                         arglag = blag.args)
            )
          }
  )

btemp1 <- do.call("rbind", btemp.locs) #joins all of the split regions together 
btemp1 %>% head()
```


Join the crossbasis results to main dataset:
```{r}
data_Brazil <- left_join(data_Brazil, btemp1)
```

Update the formula:
```{r}
mod4 = cases ~ 1 + 
  offset(log(population)) +  
  
  # long-term trends
  f(id_year, model = "rw1", hyper = hyper.iid, constr = TRUE) + 
  # seasonality
  f(id_seasonal, model = "rw2", hyper = hyper.iid, constr = TRUE) + 

  # spatial effect
  f(id_space, model = "bym2", graph='W.adj', hyper = hyper.bym, constr = TRUE) + 
  
  # meteorological linear terms
  rh + tp_sum +
  
  # temperature DLNM
  v1.l1 + v1.l2 + v1.l3 + v1.l4 + v1.l5 + 
  v2.l1 + v2.l2 + v2.l3 + v2.l4 + v2.l5 +
  v3.l1 + v3.l2 + v3.l3 + v3.l4 + v3.l5 +
  v4.l1 + v4.l2 + v4.l3 + v4.l4 + v4.l5 
```

Run `INLA`:
```{r eval = FALSE}
inla(formula = as.formula(mod4),
            data=data_Brazil,
            family="poisson",
            verbose = TRUE, 
            
            # priors fixed effects
            control.fixed = list(
              correlation.matrix = TRUE, # we add this to get the vcov, see bellow
              mean = list(rh = 0, t2m_max = 0, tp_sum = 0), 
              prec = list(rh = 0.01, t2m_max = 0.01, tp_sum = 0.01),
              prec.intercept = 0.001, mean.intercept = 0),
            
            control.compute=list(dic=TRUE, cpo=TRUE, config = TRUE), 
            control.inla=list(strategy="simplified.laplace", int.strategy="eb"),
            control.mode=list(restart = T), 
            num.threads = 1
  ) -> mod4_res # 52 seconds

saveRDS(mod4_res, file = "data/mod4.rds")
```

```{r, echo = FALSE, eval = TRUE}
mod4_res <- readRDS("data/mod4.rds")
```

Extract the DLNM:
```{r}
# Extract the fixed effects
fixedeffects <- mod4_res$summary.fixed

# Extract the variance-covariance matrix
vcov_matrix <- mod4_res$misc$lincomb.derived.covariance.matrix

#find indices of the DLNM
v_names <- grep("^v", mod4_res$names.fixed, value = TRUE)
indt <- grep("^v", mod4_res$names.fixed)

# Define temperature range for prediction
temp_grid <- quantile(data_Brazil$t2m_mean, 
                       probs = seq(0.01, 0.99, by = 0.01), 
                       na.rm = TRUE)

# the basis function on the prediction points
basis_tmean <- crossbasis(temp_grid,
                          lag = n.lags,      
                          argvar = btemp.args, 
                          arglag = blag.args)   

# and predict
predt <- crosspred(basis_tmean, coef = fixedeffects$mean[indt], vcov = vcov_matrix[indt,indt],
                   model.link = "log", bylag = 0.25, cen = mean(data_Brazil$t2m_mean))

```


Cummulative across the lags
```{r}
plot(predt, "overall")
```
Contour plot
```{r}
nlag <- n.lags

y <- predt$predvar
x <- seq(0, nlag, 0.25)
z <- t(predt$matRRfit)

library(RColorBrewer)

pal <- rev(brewer.pal(11, "PRGn"))
levels <- pretty(z, 20)
col1 <- colorRampPalette(pal[1:6])
col2 <- colorRampPalette(pal[6:11])
cols <- c(col1(sum(levels < 1)), col2(sum(levels > 1)))

filled.contour(x,y,z,
               xlab = "Lag", ylab = expression(paste("Mean Temperature (",degree,"C)")), main = "",
               col = cols,levels = levels,
               plot.axes = { axis(1, at = 0:nlag, c(0:nlag)) 
                 axis(2)})
mtext(side = 2, at = max(y)*1.1, text = "a", las = 2, cex = 1.2, line = 2)
```
```{r, eval = FALSE}
library(plotly)
p <- plot_ly()
p <- add_surface(p, x=0:12, y=-20:30, z=predt$matRRfit)
layout(p, scene = list(
  xaxis = list(title = "Lag"),
  yaxis = list(title = "Temperature", range = c(0,25)),
  zaxis = list(title = "Relative risk")
))
```

## Closing remarks

In this lab session, we have explored how to fit advanced Bayesian models using `INLA`.
We looked at linear spatially varying coefficients and distributed non-linear lag models.

We analyzed real-world data on confirmed dengue cases in Brazil from 2009 to 2024 to quantify the impact of meteorological factors. Our approach allowed the effects of temperature and rainfall to vary across space, and to exhibit delayed and non-linear patterns. We accounted for underlying spatiotemporal trends and interactions, and interpreted the results using a range of visualizations to illustrate the estimated effects.


