---
title: "Bayesian workflow and influence of priors"
subtitle: "SHARP Bayesian Modeling for Environmental Health Workshop"
author: "Robbie M. Parks"
date: "August 05 2026"
format: html
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(nimble)
library(bayesplot)
library(posterior)
library(patchwork)

theme_set(theme_minimal(base_size = 13))
color_scheme_set("blue")

set.seed(123)
options(mc.cores = 2)
```

# Goal of this computing lab session

In this lab, you will work through a **full Bayesian workflow** using `NIMBLE`, starting from simulated data so that every step is fully transparent.

By the end, you should be able to:

- simulate data from a known data-generating process;
- write Bayesian models in `NIMBLE`;
- perform **prior predictive checks**;
- diagnose **bad priors**;
- fit a model with MCMC;
- assess convergence using trace plots, \(\hat{R}\), and effective sample size;
- perform **posterior predictive checks**;
- identify when a model is misspecified; and
- improve the model and compare results across specifications.

# Define research question

Suppose we are interested in the relationship between a continuous environmental exposure and a continuous health outcome.

How do we build, check, diagnose, and refine a Bayesian model so that it can recover the underlying signal in the data?

# 0: Preparation

We will simulate data where the *true* relationship is slightly nonlinear. This is useful because it lets us see what happens when we fit both a poor model and a better model.

## Simulate data

```{r}
n <- 200

x_raw <- runif(n, -3, 3)
y_true_mean <- 2 + 1.2 * x_raw - 0.8 * x_raw^2
y <- rnorm(n, mean = y_true_mean, sd = 1)

dat <- tibble(
  x = x_raw,
  y = y,
  mu_true = y_true_mean
)

head(dat)
```

## Visualize the data

```{r}
ggplot(dat, aes(x = x, y = y)) +
  geom_point(alpha = 0.65) +
  geom_line(aes(y = mu_true), linewidth = 1) +
  labs(
    x = "Exposure",
    y = "Outcome"
  )
```

**Class question:** Looking at the plot, would a simple straight-line model be enough here?

# 1: A first model that is probably too simple

We will begin with a **linear Normal model** even though the true relationship is nonlinear.

$$
\begin{aligned}
y_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta x_i
\end{aligned}
$$

This is a useful starting point because the Bayesian workflow is not just about fitting a model once. It is about fitting, checking, diagnosing, and improving.

# 2: Prior predictive checks

A prior predictive check asks: *If our priors were true, what sort of data would the model imply before seeing the observed data?*

We will look at three cases:

1. a **bad overly wide prior**;
2. a **bad overly restrictive prior**; and
3. a more **reasonable weakly informative prior**.

## Create a grid of exposure values

```{r}
x_grid <- seq(min(dat$x), max(dat$x), length.out = 100)
```

## 2a. Bad prior: wildly too wide

```{r}
n_sims <- 200

prior_bad_wide <- map_dfr(1:n_sims, function(s) {
  alpha <- rnorm(1, 0, 2)
  beta  <- rnorm(1, 0, 10)
  sigma <- runif(1, 0, 5)

  tibble(
    sim = s,
    x = x_grid,
    y_rep = rnorm(length(x_grid), alpha + beta * x_grid, sigma)
  )
})
```

```{r}
ggplot(prior_bad_wide, aes(x = x, y = y_rep, group = sim)) +
  geom_point(alpha = 0.08) +
  # coord_cartesian(ylim = range(dat$y) + c(-10, 10)) +
  labs(
    title = "Prior predictive check: overly wide priors",
    subtitle = "These priors imply many impossible or wildly implausible datasets",
    x = "Exposure",
    y = "Simulated outcome"
  )
```

**Class question:** What looks unrealistic about these simulated datasets?

## 2b. Bad prior: too restrictive

```{r}
prior_bad_tight <- map_dfr(1:n_sims, function(s) {
  alpha <- rnorm(1, 0, 0.05)
  beta  <- rnorm(1, 0, 0.05)
  sigma <- runif(1, 0.05, 0.2)

  tibble(
    sim = s,
    x = x_grid,
    y_rep = rnorm(length(x_grid), alpha + beta * x_grid, sigma)
  )
})
```

```{r}
ggplot(prior_bad_tight, aes(x = x, y = y_rep, group = sim)) +
  coord_cartesian(ylim = range(dat$y) + c(-10, 10)) +
  geom_line(alpha = 0.08) +
  labs(
    # title = "Prior predictive check: overly restrictive priors",
    # subtitle = "These priors barely allow variation and effectively rule out meaningful effects",
    x = "Exposure",
    y = "Simulated outcome"
  )
```

**Class question:** Why is a very tight prior also a problem, even if the lines look neat?

## 2c. More reasonable prior

To make the model easier to fit, we will center the exposure.

```{r}
dat <- dat %>%
  mutate(x_c = x - mean(x))
```

```{r}
prior_reasonable <- map_dfr(1:n_sims, function(s) {
  alpha <- rnorm(1, 0, 5)
  beta  <- rnorm(1, 0, 2)
  sigma <- runif(1, 0.2, 5)

  tibble(
    sim = s,
    x_c = seq(min(dat$x_c), max(dat$x_c), length.out = 100),
    y_rep = rnorm(
      100,
      alpha + beta * seq(min(dat$x_c), max(dat$x_c), length.out = 100),
      sigma
    )
  )
})
```

```{r}
ggplot(prior_reasonable, aes(x = x_c, y = y_rep, group = sim)) +
  geom_line(alpha = 0.08) +
  labs(
    title = "Prior predictive check: more reasonable priors",
    subtitle = "These priors allow a wide range of plausible datasets without becoming absurd",
    x = "Centered exposure",
    y = "Simulated outcome"
  )
```

# 3: Write the first Bayesian model in NIMBLE

We now fit the linear model using reasonable priors.

```{r}
code_linear <- nimbleCode({
  alpha ~ dnorm(0, sd = 5)
  beta ~ dnorm(0, sd = 2)
  sigma ~ dunif(0, 10)

  for (i in 1:n) {
    mu[i] <- alpha + beta * x[i]
    y[i] ~ dnorm(mu[i], sd = sigma)
  }
})
```

## Prepare data, constants, and initial values

```{r}
constants_linear <- list(n = nrow(dat))
data_linear <- list(y = dat$y, x = dat$x_c)
inits_linear <- list(alpha = 0, beta = 0, sigma = 1)
```

## Build model and inspect MCMC configuration

```{r}
model_linear <- nimbleModel(
  code_linear,
  constants = constants_linear,
  data = data_linear,
  inits = inits_linear
)

conf_linear <- configureMCMC(model_linear)
conf_linear$printSamplers()
```

# 4: First MCMC run: intentionally too short

We will deliberately run a short MCMC first to mimic a common classroom situation where a model has been fit, but diagnostics have not really been checked yet.

```{r}
samples_linear_short <- nimbleMCMC(
  code = code_linear,
  constants = constants_linear,
  data = data_linear,
  inits = inits_linear,
  niter = 300,
  nburnin = 250,
  nchains = 2,
  setSeed = c(1,2),
  samplesAsCodaMCMC = TRUE
)
```

## Posterior summaries

```{r}
summarise_draws(
  samples_linear_short,
  default_summary_measures()
)
```

## Convergence diagnostics

```{r}
summarise_draws(
  samples_linear_short,
  default_convergence_measures()
)
```

## Trace plots

```{r}
mcmc_trace(samples_linear_short)
```

**Class question:** Do these chains look fully stable and well mixed?

# 5: Better MCMC run

Now we run the same model more carefully.

```{r}
samples_linear <- nimbleMCMC(
  code = code_linear,
  constants = constants_linear,
  data = data_linear,
  inits = inits_linear,
  niter = 12000,
  nburnin = 2000,
  thin = 5,
  nchains = 3,
  setSeed = c(1:3),
  samplesAsCodaMCMC = TRUE
)
```

## Posterior summaries

```{r}
summarise_draws(
  samples_linear,
  default_summary_measures()
)
```

## Convergence diagnostics

```{r}
summarise_draws(
  samples_linear,
  default_convergence_measures()
)
```

## Trace plots

```{r}
mcmc_trace(samples_linear)
```

## Posterior intervals

```{r}
mcmc_intervals(samples_linear, pars = c("alpha", "beta", "sigma"))
```

**Class question:** Which parameter would represent the shape of the fit here?

# 6: Posterior predictive checks for the linear model

A posterior predictive check asks: *If the fitted model were true, could it generate data that look like what we actually observed?*

## Extract posterior draws

```{r}
draws_linear <- as.matrix(samples_linear)

dim(draws_linear)
```

## Simulate replicated outcomes from the posterior

```{r}
n_draws_ppc <- 300
draw_ids <- sample(1:nrow(draws_linear), n_draws_ppc)

yrep_linear <- matrix(NA_real_, nrow = n_draws_ppc, ncol = nrow(dat))

for (j in seq_along(draw_ids)) {
  d <- draw_ids[j]
  mu_j <- draws_linear[d, "alpha"] + draws_linear[d, "beta"] * dat$x_c
  sigma_j <- draws_linear[d, "sigma"]
  yrep_linear[j, ] <- rnorm(nrow(dat), mean = mu_j, sd = sigma_j)
}
```

## Compare observed and replicated distributions

```{r}
ppc_dens_overlay(y = dat$y, yrep = yrep_linear[1:50, ])
```

## Compare means and standard deviations

```{r}
ppc_stat(y = dat$y, yrep = yrep_linear, stat = "mean")
```

```{r}
ppc_stat(y = dat$y, yrep = yrep_linear, stat = "sd")
```

## Compare fitted relationship with observed data

```{r}
pp_linears <- map_dfr(1:100, function(j) {
  d <- draw_ids[j]
  tibble(
    draw = j,
    x_c = dat$x_c,
    mu = draws_linear[d, "alpha"] + draws_linear[d, "beta"] * dat$x_c
  )
})

ggplot(dat, aes(x = x_c, y = y)) +
  geom_point(alpha = 0.55) +
  geom_line(
    data = pp_linears,
    aes(y = mu, group = draw),
    alpha = 0.08,
    color = "steelblue"
  ) +
  labs(
    title = "Posterior mean functions from the linear model",
    subtitle = "The model can only fit straight lines",
    x = "Centered exposure",
    y = "Outcome"
  )
```

**Class question:** What important feature of the data is this model failing to capture?

# 7: A better model: allow curvature

Because the true relationship is nonlinear, we now fit a quadratic mean structure.

$$
\begin{aligned}
y_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_1 x_i + \beta_2 x_i^2
\end{aligned}
$$

## Add a squared term

```{r}
dat <- dat %>%
  mutate(x2 = x_c^2)
```

## Prior predictive check for the quadratic model

```{r}
prior_quad <- map_dfr(1:n_sims, function(s) {
  alpha <- rnorm(1, 0, 5)
  beta1 <- rnorm(1, 0, 2)
  beta2 <- rnorm(1, 0, 1)
  sigma <- runif(1, 0.2, 5)

  x_seq <- seq(min(dat$x_c), max(dat$x_c), length.out = 100)

  tibble(
    sim = s,
    x_c = x_seq,
    y_rep = rnorm(100, alpha + beta1 * x_seq + beta2 * x_seq^2, sigma)
  )
})
```

```{r}
ggplot(prior_quad, aes(x = x_c, y = y_rep, group = sim)) +
  geom_line(alpha = 0.06) +
  labs(
    title = "Prior predictive check for the quadratic model",
    subtitle = "The model now allows both linear and curved relationships",
    x = "Centered exposure",
    y = "Simulated outcome"
  )
```

## Write the quadratic NIMBLE model

```{r}
code_quad <- nimbleCode({
  alpha ~ dnorm(0, sd = 5)
  beta1 ~ dnorm(0, sd = 2)
  beta2 ~ dnorm(0, sd = 1)
  sigma ~ dunif(0, 10)

  for (i in 1:n) {
    mu[i] <- alpha + beta1 * x[i] + beta2 * x2[i]
    y[i] ~ dnorm(mu[i], sd = sigma)
  }
})
```

## Prepare inputs

```{r}
constants_quad <- list(n = nrow(dat))
data_quad <- list(y = dat$y, x = dat$x_c, x2 = dat$x2)
inits_quad <- list(alpha = 0, beta1 = 0, beta2 = 0, sigma = 1)
```

## Fit the quadratic model

```{r}
samples_quad <- nimbleMCMC(
  code = code_quad,
  constants = constants_quad,
  data = data_quad,
  inits = inits_quad,
  niter = 12000,
  nburnin = 2000,
  thin = 5,
  nchains = 3,
  setSeed = c(1:3),
  samplesAsCodaMCMC = TRUE
)
```

## Summaries and diagnostics

```{r}
summarise_draws(
  samples_quad,
  default_summary_measures()
)
```

```{r}
summarise_draws(
  samples_quad,
  default_convergence_measures()
)
```

```{r}
mcmc_trace(samples_quad)
```

# 8: Posterior predictive checks for the quadratic model

## Extract posterior draws

```{r}
draws_quad <- as.matrix(samples_quad)
```

## Simulate replicated outcomes

```{r}
draw_ids_quad <- sample(1:nrow(draws_quad), n_draws_ppc)

yrep_quad <- matrix(NA_real_, nrow = n_draws_ppc, ncol = nrow(dat))

for (j in seq_along(draw_ids_quad)) {
  d <- draw_ids_quad[j]
  mu_j <- draws_quad[d, "alpha"] +
    draws_quad[d, "beta1"] * dat$x_c +
    draws_quad[d, "beta2"] * dat$x2
  sigma_j <- draws_quad[d, "sigma"]
  yrep_quad[j, ] <- rnorm(nrow(dat), mean = mu_j, sd = sigma_j)
}
```

## Compare distributions

```{r}
ppc_dens_overlay(y = dat$y, yrep = yrep_quad[1:50, ])
```

## Compare summary statistics

```{r}
ppc_stat(y = dat$y, yrep = yrep_quad, stat = "mean")
```

```{r}
ppc_stat(y = dat$y, yrep = yrep_quad, stat = "sd")
```

## Compare fitted curves with observed data

```{r}
pp_quads <- map_dfr(1:100, function(j) {
  d <- draw_ids_quad[j]
  tibble(
    draw = j,
    x_c = dat$x_c,
    mu = draws_quad[d, "alpha"] +
      draws_quad[d, "beta1"] * dat$x_c +
      draws_quad[d, "beta2"] * dat$x2
  )
})

ggplot(dat, aes(x = x_c, y = y)) +
  geom_point(alpha = 0.55) +
  geom_line(
    data = pp_quads,
    aes(y = mu, group = draw),
    alpha = 0.08,
    color = "darkgreen"
  ) +
  labs(
    title = "Posterior mean functions from the quadratic model",
    subtitle = "The model now captures the curvature in the data much better",
    x = "Centered exposure",
    y = "Outcome"
  )
```

**Class question:** Compared with the linear model, what improved here?

# 9: Compare the poor model and the improved model

## Compare posterior predictive distributions side by side

```{r}
p1 <- ppc_dens_overlay(y = dat$y, yrep = yrep_linear[1:50, ]) +
  ggtitle("Linear model")

p2 <- ppc_dens_overlay(y = dat$y, yrep = yrep_quad[1:50, ]) +
  ggtitle("Quadratic model")

p1 + p2
```

## Compare fitted mean curves
```{r}
linear_fits <- sapply(1:200, function(j) {
  d <- draw_ids[j]
  draws_linear[d, "alpha"] + draws_linear[d, "beta"] * dat$x_c
})

quad_fits <- sapply(1:200, function(j) {
  d <- draw_ids_quad[j]
  draws_quad[d, "alpha"] +
    draws_quad[d, "beta1"] * dat$x_c +
    draws_quad[d, "beta2"] * dat$x2
})

linear_mean_curve <- tibble(
  x_c = dat$x_c,
  mu = rowMeans(linear_fits)
) %>%
  arrange(x_c)

quad_mean_curve <- tibble(
  x_c = dat$x_c,
  mu = rowMeans(quad_fits)
) %>%
  arrange(x_c)

ggplot(dat, aes(x = x_c, y = y)) +
  geom_point(alpha = 0.45) +
  geom_line(
    data = linear_mean_curve,
    aes(x = x_c, y = mu),
    color = "steelblue",
    linewidth = 1
  ) +
  geom_line(
    data = quad_mean_curve,
    aes(x = x_c, y = mu),
    color = "darkgreen",
    linewidth = 1
  ) +
  labs(
    title = "Posterior mean fit comparison",
    subtitle = "Blue = linear model, green = quadratic model",
    x = "Centered exposure",
    y = "Outcome"
  )
```

# 10: Sensitivity analysis

The next step in a Bayesian workflow is to ask whether our conclusions are sensitive to reasonable alternative priors.

Here we keep the quadratic model but make the prior on the curvature term a little wider.

```{r}
code_quad_sens <- nimbleCode({
  alpha ~ dnorm(0, sd = 5)
  beta1 ~ dnorm(0, sd = 2)
  beta2 ~ dnorm(0, sd = 2)
  sigma ~ dunif(0, 10)

  for (i in 1:n) {
    mu[i] <- alpha + beta1 * x[i] + beta2 * x2[i]
    y[i] ~ dnorm(mu[i], sd = sigma)
  }
})
```

```{r}
samples_quad_sens <- nimbleMCMC(
  code = code_quad_sens,
  constants = constants_quad,
  data = data_quad,
  inits = inits_quad,
  niter = 12000,
  nburnin = 2000,
  thin = 5,
  nchains = 3,
  setSeed = c(1:3),
  samplesAsCodaMCMC = TRUE
)
```

## Compare posterior intervals for the key parameters

```{r}
mcmc_intervals(
  samples_quad,
  pars = c("beta1", "beta2")
)
```

```{r}
mcmc_intervals(
  samples_quad_sens,
  pars = c("beta1", "beta2")
)
```
